OPTIMAL PRICING AND INFORMAL SHARING: \\ EVIDENCE FROM PIPED WATER IN MANILA

### 2018-12-06 phil_paper/jmp_wjv.tex
fixing stuff

	Public utilities subsidize fixed connection fees with high marginal prices to provide access especially for the poor, but this policy can have the opposite effect when many households share connections.  Using data on 1.5 million water connections in Manila, I establish that informal sharing networks provide \input{tables/shr_use_from_neighbor}\unskip\% 
	of total access.  I structurally estimate household water demand across three sources: purchasing directly from the provider, sharing with a neighbor's tap, or buying from a small-scale vendor.  The model predicts that low fixed fees and high marginal prices decrease access by weakening incentives to share water.  In contrast, the optimal pricing policy features a high fixed fee and low marginal price, increases shared connections, and ensures nearly universal access to piped water compared to current prices in Manila.

@@

\section{Introduction}
 % \footnote{\cite{laffont2005regulation} emphasizes high costs of public funds in developing countries (with a deadweight loss of 2.5 for the Philippines) due to limited ability to enter long-term contracts with government, credit constraints, inefficient taxation, and corruption.}
 % To ensure widespread access while covering costs, a 
Improving access to piped water can generate substantial health benefits\footnote{\cite{zhang2016long,kosec2014child,gamper2010impact}, and \cite{galiani2005water} focus on health improvements in developing countries while \cite{cutler2005role} look historically at the US.} as well as improvements in household well-being through time savings and reduced intra-household conflict.\footnote{\cite{devoto2012happiness} provide water through a randomized-controlled trial in Morocco finding few health effects but large benefits to household welfare.}  However, the share of urban households without piped water has remained stable at 5\% for middle income countries and at 20\% for low income countries (World Bank [2015]).  While external health and other benefits may merit subsidizing piped water, government subsidies are often limited by high costs of public funds and risk of regulatory capture in these countries (\cite{laffont2005regulation}).  In many cases, prices must be set to fully recover costs.  A popular solution is to subsidize upfront connection fees with high marginal prices for monthly usage.\footnote{\cite{mcintosh2003asian} and \cite{komives2006distributional} from the World Bank and Asian Development Bank argue for low connection fees.  Using a randomized controlled trial from Morocco, \cite{devoto2012happiness} confirm that modest assistance with connection fees can lead to increases in new water connections.}  Over 70\% of developing cities also increase marginal prices with monthly usage so that large users cross-subsidize access and usage for smaller, poorer users.\footnote{This strategy is also commonly referred to as ``Increasing Block Tariffs'' or  ``Convex Pricing'' (\cite{borenstein2012redistributional}). \cite{hoque2013state} and \cite{boland2000political} provide overviews of the use of increasing marginal prices in the water sector in developing countries.  The statistic comes from author calculations with data kindly provided by \cite{hoque2013state}.}  These pricing approaches have been shown to work well when each water connection serves a single household.  Recent empirical papers in South Africa and C{\^o}te d'Ivoire predict gains in both access and total welfare from increasing marginal prices (\cite{szabo2015value,diakite2009proposal}).  However, many households in developing cities share connections with their neighbors, which can have large implications for water pricing.

Land-tenure issues, credit-constraints, and administrative costs often prevent disadvantaged households from purchasing their own connections (\cite{devoto2012happiness}).  Also, many cities maintain high upfront fees despite policy recommendations otherwise (\cite{hoque2013state}).  As a result, 70\% of African utilities report frequent water sharing between households (\cite{keener2010provision}) ranging from fetching from neighbors with hoses and buckets in Tangiers, Morocco (\cite{devoto2012happiness}) to drawing from single taps that each serve compounds in Kumasi, Ghana (\cite{whittington1992possible}).\footnote{\cite{nauges2006water} finds further evidence of widespread water sharing in Sri Lanka.}  While documenting the prevalence of informal sharing, previous research has not yet examined the importance of informal sharing for optimal pricing policy.

This paper analyzes the welfare effects of pricing policies in the presence of informal water sharing.  I build a model of household water demand that includes both (1) the upfront choice of whether to purchase a connection, share with a neighbor, or use non-piped sources as well as (2) monthly consumption choices.  I also collect new survey evidence on informal sharing contracts to inform several features of the model design.  The structural estimation leverages administrative data from 1.5 million water connections in Manila from 2010 to 2015, which link to additional survey data on households sharing connections.

These estimates allow me to solve a standard regulator pricing problem: setting prices to cover fixed and marginal costs while maximizing consumer surplus.  I also let the regulator place extra weight on surplus for smaller, poorer users.  In the presence of sharing, the standard approach --- low fixed fees and high marginal prices --- excludes many small-users from piped water by reducing incentives to share connections.  Instead, these households substitute toward non-piped water sources, which have lower marginal prices, but varying quality.\footnote{Common non-piped water sources in Manila include deepwells, tanker trucks, and water refilling stations.  The model collapses all non-piped sources into a single ``vendor'' option.}  The opposite approach --- high fixed fees and low marginal prices --- maximizes both access and welfare even for smaller users.  High fixed fees are paid in full by households who connect individually and are split between sharing households.  Informal sharing arrangements allow small-users to access piped water from their neighbors for even lower marginal prices than those offered by non-piped sources.  Shared connections also let the city save on installation and maintenance costs, providing an efficient means for extending piped water to low-demand households.\footnote{Cost savings are passed on to consumers through lower prices.}

The welfare impacts of these pricing policies depend not only on household-level demand, but also on the relative costs of sharing water; these may include both marginal costs (ie. time spent fetching from neighbors or reduced pressure from re-piping water) as well as savings from splitting the fixed costs of a connection.  Measuring demand and sharing costs presents two central challenges: how can we recover household-level demand when we only observe water consumption for each connection, and how can we separately identify the costs of sharing from preferences for water?  The empirical approach addresses these challenges in three steps.

The first step builds a tractable model of water sharing to estimate household-level demand from monthly usage data for each connection.  Informed by a new survey, this model embeds how (1) households face challenges in observing each other's monthly demand and (2) households often split the monthly water bill evenly.  These assumptions on sharing behavior ensure that connection-level demand can be expressed as a sum of underlying household demands even in the presence of non-linear prices.  I use variation in the non-linear price schedule (both across and within connections) to identify water demand for each connection before disaggregating preferences to the household-level.  This approach extends previous models of optimization over non-linear prices by including sharing and allowing households to anticipate unobserved shocks to their demand (\cite{moffitt1986econometrics,burtless1978effect}).

To estimate the marginal costs of sharing water, the second step analyzes a quasi-experiment where breaks in service pipes force some households to switch from their own water connections to neighbors' connections.  I then measure how much less water households use from their neighbors' connections and attribute this gap to the marginal costs of sharing.  This exercise recovers sizable marginal costs on the order of \input{tables/hassle_percentage}\unskip\% of the average marginal water price.\footnote{This revealed-preference approach builds on a descriptive literature that identifies potential sources of water sharing frictions (\cite{whittington1992possible,nauges2006water}).}

The third step maps the share of households using from each source --- connecting alone, sharing with a neighbor, or using from a non-piped vendor --- into an estimate of each source's relative fixed cost.  The empirical approach leverages the assumption that given water preferences (from step one) and marginal sharing costs (from step two), any remaining variation in water source choices must be due to differences in their fixed costs.  Since sharing decisions are strategic in nature, I specify a simple game where households sequentially choose their sources and agree to split any surplus from sharing.  This revealed-preference approach to estimating fixed costs is especially relevant in a developing context because on top of the initial connection and monthly service fees, households often face fixed costs that are difficult to observe, including permitting fees, plumbing maintenance, as well as land-tenure issues in qualifying for a water connection.  In fact, I estimate that the connection and service fees account for only around half of the total fixed costs facing households.  This structural approach complements previous experimental evidence of fixed costs presented in \cite{devoto2012happiness}.  By considering both source and quantity choices, this model also expands on the demand estimation literature for public utilities that has so far focused almost exclusively on monthly quantity choices (\cite{diakite2009proposal,mcrae2014infrastructure,olmstead2009reduced,szabo2015value}).

@@

With these estimates, I simulate a series of counterfactual policies.  First, I consider a connection fee subsidy policy consistent with recommendations from the World Bank, NGOs, and other researchers to improve water access for the poor (\cite{komives2005water,jimenez2014factors,mcintosh2003asian}).  I discount the fixed fee by 12\% while increasing the tariff schedule proportionally to ensuring that the water provider exactly covers costs.  This policy actually reduces access by incentivizing households to substitute toward non-piped vendors.  Next, I compute an optimal tariff schedule to maximize consumer surplus while exactly covering production costs.  Compared to the current tariff, I find that a simple, two-part tariff characterized by a fixed fee and low, flat marginal price can produce gains on the order of \input{tables/tpt_improvement_percent_sharing_group}\unskip\% of consumer surplus.\footnote{With quasi-linear utility, consumer surplus for a given tariff is calculated as the difference between household utility under this tariff and household utility from using the non-piped, vendor option.}  These gains are driven both by pricing closer to marginal cost as well as by increasing sharing, which reduces the number of connections to install and maintain.  Finally, I compare the optimal tariff with a social tariff that maximizes consumer surplus only for the bottom 50\% of users and is intended to capture goals of ensuring minimum usage and access as well as income redistribution.  With sharing, social tariffs are ineffective at reallocating surplus from large to small users; however in a counterfactual without sharing, social tariffs achieve substantial redistribution.  These findings suggest that informal sharing networks constrain the extent to which non-linear prices can redistribute surplus.

@@

This paper demonstrates that pricing policies intended to increase water access especially among the poor can have unintended consequences when low-income households can simply access the piped water network through their neighbors.

The counterfactual exercises find that a simple two-part tariff structure outperforms the increasing block tariff currently implemented in Manila in terms of overall access to water, total welfare, and even welfare among the bottom half of water users.  In this counterfactual, high fixed costs of accessing an individual connection allow for sharing networks to play an important role in extending water access despite sizable hassle costs.  I also find that complex non-linear price schedules are largely ineffective in terms of achieving both efficiency and equity goals when sharing networks are readily available.

Taken together, these results suggest that households may act as efficient subcontractors in redistributing water to their neighbors.  Instead of sending trained engineers to install and maintain connections in dense neighborhoods, providers can instead rely on informal networks to extend water distribution.  From this perspective, cities may benefit from considering how prices not only might encourage access and redistribute consumer surplus, but also may affect the endogenous formation of water sharing networks.  For example, price discriminating according to neighborhood income instead of by monthly usage levels may achieve similar goals of redistribution without distorting sharing networks.  While this policy may not be appropriate in Manila since neighborhoods are often very heterogeneous in income, this approach may work well in other contexts where neighborhoods are more segregated, such as in Colombia where electricity rates vary according to the social category of each neighborhood (\cite{mcrae2014infrastructure}).

@@

In focusing on water sharing behavior, this approach abstracts away from several features that may also affect optimal water pricing.  First, I do not directly model environmental, health, or other externalities associated with piped water use.  Water shortages may lead policymakers to promote conservation through higher marginal prices while positive health externalities may motivate further subsidizing piped water (\cite{timmins2002measuring}, \cite{galiani2005water}).\footnote{\cite{bennett2012does} shows that piped water can also lead to negative externalities through worse sanitation, possibly motivating higher prices for piped water in some cases.}  Second, I assume that government regulator is able to fully observe production costs for the water provider, eliminating scope for prices to correct for possible principal-agent frictions as characterized by \cite{laffont1993theory}.  Finally, the counterfactual exercises simulate the public-private partnership regulatory framework in Manila by assuming that revenue must fully cover production costs.  Additional data on the costs of raising public funds and risks of regulatory capture would allow for a more flexible analysis taking into account the possibility of using public funding for water infrastructure (\cite{laffont2005regulation}).

>>>>

	In developing cities, high costs of connecting to piped water lead many households to use low-quality, non-piped sources.  In response, cities often set low fixed prices so that more households connect and high marginal prices to cover production costs.  This paper identifies a common alternative source --- sharing a neighbor's connection --- that has large implications for these pricing strategies.  With microdata from 1.5 million water connections in Manila, I structurally estimate water demand in the presence of water sharing and solve the city's pricing problem.  I find that high marginal prices weaken incentives to share water between neighbors.  In contrast, the optimal pricing policy features high fixed and low marginal prices, increases shared connections, and ensures nearly universal access to piped water compared to current prices in Manila.

% 	However, I find that these prices

% I find that

% 	In particular, I find that high marginal prices weaken incentives to share connections,

% 	access and welfare by weakening incentives to share water.

% 	or share with their neighbors' connections.

% 	.  I then analyze the regulator's pricing problem: choosing fixed and marginal prices to maximize welfare while covering costs.

% 	With these estimates, I find that the presence of sharing reverses

% 	At the same time, cities must choose relative fixed and marginal prices cover production costs by ,

% 	which have large implications for piped water access and usage.  Cities often subsidize

% 	 low fixed and high marginal prices to offset other connection costs and provide access especially for the poor; however, I find that

% 	often leading them to use from low-quality vendors or share with their neighbors' connections.

% 	In developing cities, households face many costs of connecting to piped water, often leading them to use from low-quality vendors or share with their neighbors' connections.  At the same time, many cities must set prices to cover production costs.

% 	Many cities respond by adjusting fixed and marginal prices to ensure access while covering costs.

% 	To ensure piped water access, cities are often constrained to adjusting fixed and marginal prices while covering production costs.

% 	The primary way cities can respond is by  setting fixed and marginal prices still

% 	respond by adjusting the fixed and marginal prices

% 	Since water utilities must often cover their costs, cities are constrained to

% 	With data from 1.5 million water connections in Manila, I structurally estimate household water demand to analyze optimal fixed and marginal prices for piped water in the presence of water sharing.  Cities often set low fixed and high marginal prices to offset other connection costs and provide access especially for the poor; however, I find that

% 	To offset connection costs, cities often subsidize the fixed price with high marginal prices

% 	 to provide access especially for the poor; however,

% 	Using data on 1.5 million water connections in Manila, I structurally estimate household water demand across three sources: purchasing directly from the provider, sharing with a neighbor's tap, or buying from a small-scale vendor.

% to maximize welfare while covering their costs

% 	This paper analyzes how

% 	To offset these connection costs, cities often subsidize their fixed connection fees with high marginal prices to provide access especially for the poor; however,

% 	Public utilities choose fixed and marginal prices to recover costs.

% 	This paper analyzes the welfare effects of pricing policies for piped water in the presence of water sharing.

% 	Public utilities subsidize fixed connection fees with high marginal prices to provide access especially for the poor.

% 	 but this policy can have the opposite effect when many households share connections.

% 	The model predicts that low fixed fees and high marginal prices decrease access and welfare by weakening incentives to share water.

% 	In contrast, the optimal pricing policy features a high fixed fee and low marginal price, increases shared connections, and ensures nearly universal access to piped water compared to current prices in Manila.

% 	I establish that informal sharing networks provide \input{tables/shr_use_from_neighbor}\unskip\% 
% 	of total piped water access.

@@

\section{Introduction}
 % \footnote{\cite{laffont2005regulation} emphasizes high costs of public funds in developing countries (with a deadweight loss of 2.5 for the Philippines) due to limited ability to enter long-term contracts with government, credit constraints, inefficient taxation, and corruption.}
 % To ensure widespread access while covering costs, a 
Improving access to piped water can generate substantial health benefits\footnote{\cite{zhang2016long,kosec2014child,gamper2010impact}, and \cite{galiani2005water} focus on health improvements in developing countries while \cite{cutler2005role} look historically at the US.} as well as improvements in household well-being through time savings and reduced intra-household conflict.\footnote{\cite{devoto2012happiness} provide water through a randomized-controlled trial in Morocco finding few health effects but large benefits to household welfare.}  Yet, the share of urban households without piped water has remained stable at 5\% for middle income countries and at 20\% for low income countries (World Bank [2015]).  While external health and other benefits may merit subsidizing piped water, any subsidies are often limited by high costs of public funds and risk of regulatory capture in developing countries (\cite{laffont2005regulation}).  In many cases, prices must be set to fully recover costs.  To ensure access to piped water while covering costs, cities often subsidize upfront connection fees with high marginal prices for monthly usage.\footnote{\cite{mcintosh2003asian} and \cite{komives2006distributional} from the World Bank and Asian Development Bank argue for low connection fees.  Using a randomized controlled trial from Morocco, \cite{devoto2012happiness} confirm that modest assistance with connection fees can lead to increases in new water connections.}  Over 70\% of developing cities increase marginal prices with monthly usage so that large users cross-subsidize access and usage for smaller, poorer users.\footnote{This strategy is also commonly referred to as ``Increasing Block Tariffs'' or  ``Convex Pricing'' (\cite{borenstein2012redistributional}). \cite{hoque2013state} and \cite{boland2000political} provide overviews of the use of increasing marginal prices in the water sector in developing countries.  The statistic comes from author calculations with data kindly provided by \cite{hoque2013state}.}  These pricing approaches have been shown to work well when each water connection serves a single household.  Recent empirical papers in South Africa and C{\^o}te d'Ivoire predict gains in both access and total welfare from increasing marginal prices (\cite{szabo2015value,diakite2009proposal}).  However, many households in developing cities share connections with their neighbors, which can have large implications for water pricing.

% often increase the costs of purchasing a water connection particularly for disadvantaged households

Insecure land-tenure, credit-constraints, and administrative costs often make it difficult for disadvantaged households to purchase water connections (\cite{devoto2012happiness}).  Also, many cities maintain high upfront fees despite policy recommendations otherwise (\cite{hoque2013state}).  As a result, 70\% of African utilities report frequent water sharing between households (\cite{keener2010provision}) ranging from fetching from neighbors with hoses and buckets in Tangiers, Morocco (\cite{devoto2012happiness}) to drawing from single taps that each serve compounds in Kumasi, Ghana (\cite{whittington1992possible}).\footnote{\cite{nauges2006water} finds further evidence of widespread water sharing in Sri Lanka.}  While documenting the prevalence of these  sharing networks, research has not yet examined the importance of water sharing for optimal pricing policy.

This paper integrates sharing behavior into water demand in order to analyze the welfare effects of pricing policies.  I build a model of household water demand that includes both (1) the upfront choice of whether to purchase a connection, share with a neighbor, or use non-piped sources as well as (2) monthly consumption choices.  I also collect new survey evidence on sharing contracts to inform several features of the model design.  The structural estimation leverages administrative data from 1.5 million water connections in Manila from 2010 to 2015, which link to additional survey data on households sharing connections.

These estimates allow me to solve a standard regulator pricing problem: setting prices to cover fixed and marginal costs while maximizing consumer surplus.  I also let the regulator place extra weight on surplus for smaller, poorer users.  In the presence of sharing, the standard approach --- low fixed fees and high marginal prices --- excludes many small-users from piped water by reducing incentives to share connections.  Instead, these households substitute toward non-piped water sources, which have lower marginal prices, but varying quality.\footnote{Common non-piped water sources in Manila include deepwells, tanker trucks, and water refilling stations.}  The opposite approach --- high fixed fees and low marginal prices --- maximizes both access and welfare even for smaller users.  High fixed fees are paid in full by households who connect individually and are split between sharing households.  Informal sharing arrangements enable small-users to access piped water from their neighbors for lower marginal prices than those offered by non-piped sources.  Shared connections also let the city save on installation and maintenance costs, providing an efficient means for extending piped water to low-demand households.\footnote{Cost savings are passed on to consumers through lower prices.}

The welfare impacts of these pricing policies depend not only on household-level demand, but also on the relative costs of sharing water; these may include both marginal costs (ie. time spent fetching from neighbors or reduced pressure from re-piping water) as well as savings from splitting the fixed costs of a connection.  Measuring these features presents two central challenges: how can we recover household-level demand when we only observe water consumption for each connection, and how can we separately identify the costs of sharing and preferences for water?  The empirical approach addresses these challenges in three steps.

The first step builds a tractable model of water sharing to estimate household-level demand from monthly usage data for each connection.  Informed by new survey evidence, this model embeds how (1) households face challenges in observing each other's monthly demand and (2) households often split the monthly water bill evenly.  These assumptions on sharing behavior ensure that connection-level demand can be expressed as a sum of underlying household demands even in the presence of non-linear prices.  I use variation in the non-linear price schedule (both across and within connections) to identify water demand for each connection before disaggregating preferences to the household-level.  This approach extends previous models of optimization over non-linear prices by including sharing and allowing households to anticipate unobserved shocks to their demand (\cite{moffitt1986econometrics,burtless1978effect}).

To estimate the marginal costs of sharing water, the second step analyzes a quasi-experiment where breaks in service pipes force some households to switch from their own water connections to their neighbors' connections.  I then measure how much less water households use from their neighbors' connections and attribute this gap to the marginal costs of sharing.  This exercise recovers sizable marginal costs on the order of \input{tables/hassle_percentage}\unskip\% of the average marginal water price.\footnote{This revealed-preference approach builds on a descriptive literature that identifies potential sources of water sharing frictions (\cite{whittington1992possible,nauges2006water}).}

The third step maps the share of households using from each source --- connecting alone, sharing with a neighbor, or using from a non-piped vendor --- into an estimate of each source's relative fixed cost to consumers.  The empirical approach leverages the assumption that given water preferences (from step one) and marginal sharing costs (from step two), any remaining variation in water source choices must be due to differences in their fixed costs.  Since sharing decisions are strategic in nature, I specify a simple game where households sequentially choose their sources and agree to split any surplus from sharing.

This revealed-preference approach to estimating fixed costs is especially relevant in a developing context because on top of the initial connection and monthly service fees, households often face other fixed costs that are difficult to observe, including permitting fees, plumbing maintenance, as well as land-tenure issues in qualifying for a water connection.  In fact, I estimate that the connection and service fees account for only around half of the total fixed costs.  This structural approach complements previous experimental evidence of fixed costs by \cite{devoto2012happiness}.  By considering both source and quantity choices, this model also expands on the demand estimation literature for public utilities that has so far focused almost exclusively on monthly quantity choices (\cite{diakite2009proposal,mcrae2014infrastructure,olmstead2009reduced,szabo2015value}).

@@

With these estimates, I simulate a series of counterfactual policies.  First, I consider a connection fee subsidy policy consistent with recommendations from the World Bank, NGOs, and other researchers to improve water access for the poor (\cite{komives2005water,jimenez2014factors,mcintosh2003asian}).  I discount the fixed fee by 12\% while increasing the tariff schedule proportionally to ensuring that the water provider exactly covers costs.  This policy actually reduces access by incentivizing households to substitute toward non-piped vendors.  Next, I compute an optimal tariff schedule to maximize consumer surplus while exactly covering production costs.  Compared to the current tariff, I find that a simple, two-part tariff characterized by a fixed fee and low, flat marginal price can produce gains on the order of \input{tables/tpt_improvement_percent_sharing_group}\unskip\% of consumer surplus.\footnote{With quasi-linear utility, consumer surplus for a given tariff is calculated as the difference between household utility under this tariff and household utility from using the non-piped, vendor option.}  These gains are driven both by pricing closer to marginal cost as well as by increasing sharing, which reduces the number of connections installed and maintained by the utility.

Finally, I compute a tariff that maximizes consumer surplus only for the bottom 50\% of users and is intended to capture goals of ensuring minimum usage and access as well as income redistribution.  With sharing, this tariff is ineffective at reallocating surplus from large to small users; however in a counterfactual without sharing, this tariff achieves substantial redistribution.  These findings suggest that informal sharing networks constrain the extent to which non-linear prices can redistribute surplus.

@@

Observing that ``connection fees are too high, and lump sum upfront payments exclude the poor,'' the Asian Development Bank has argued for steeply increasing water prices and has emphasized that ``the irony of the situation is that the main way to help the poor is to substantially raise tariffs.''\footnote{This quote is from \cite{mcintosh2003asian} pages 40 \& 45.  The authors advocate for a ``three-block system with the lifeline consumption rate at 0-6 m3 (tariff limited to 5\% of household income for the lowest 10th percentile), consumption of 6-20 m3 charged at a rate to recover all financial costs, and consumption of more than 20 m3 charged at a penalty rate equal to about \$1/m3 (to conserve water)'' (\cite{mcintosh2003asian} p.76).}  In this paper, I show that the presence of informal sharing reverses this common intuition among policymakers: high-fixed fees no longer act as a barrier for low-income households when they can split these fees by sharing with their neighbors.  Moreover, low marginal prices help to prevent low-income households from substituting toward non-piped, vendor sources.

By capturing demand heterogeneity, the costs of sharing, as well as source choices, the structural model provides additional insights into how a range of non-linear pricing policies affect both access and welfare.  In the presence of sharing, the model finds that complex non-linear pricing is largely ineffective in achieving either efficiency or equity; sharing provides a resale market, undermining any goals of price discrimination with non-linear pricing.  The structural model also allows me to forecast how sharing costs affect optimal prices.  I find that optimal prices balance competing goals of pricing at marginal cost and taxing sharing when it is costly.

In focusing on water sharing behavior, this approach abstracts away from several features that may also affect optimal water pricing.  First, I do not directly model environmental, health, or other externalities associated with piped water use.  Water shortages may lead policymakers to promote conservation through higher marginal prices while positive health externalities may motivate further subsidizing piped water (\cite{timmins2002measuring}, \cite{galiani2005water}).\footnote{\cite{bennett2012does} shows that piped water can also lead to negative externalities through worse sanitation, possibly motivating higher prices for piped water in some cases.}  Second, I assume that the government regulator is able to fully observe production costs for the water provider, eliminating scope for prices to correct for possible principal-agent frictions as characterized by \cite{laffont1993theory}.  Finally, the counterfactual exercises simulate the public-private partnership regulatory framework in Manila by assuming that revenue must fully cover production costs.  Additional data on the costs of raising public funds and risks of regulatory capture would allow for a more flexible analysis taking into account the possibility of using public subsidies for water infrastructure (\cite{laffont2005regulation}).

Taken together, this paper suggests that households may act as efficient subcontractors in redistributing water to their neighbors.  Instead of sending trained engineers to install and maintain connections in dense neighborhoods, providers can instead rely on informal networks to extend water distribution.  From this perspective, cities may benefit from considering how prices not only might encourage access and redistribute consumer surplus, but also may affect the endogenous formation of water sharing networks.  For example, price discriminating according to neighborhood income instead of by monthly usage levels may achieve similar goals of redistribution without distorting sharing networks.  While this policy may not be appropriate in Manila since neighborhoods are often very heterogeneous in income, this approach may work well in other contexts where neighborhoods are more segregated, such as in Colombia where electricity rates vary according to the social category of each neighborhood (\cite{mcrae2014infrastructure}).

@@

%On top of the broad recommendation for high fixed fees and low marginal prices, the structural model also

% lowering fixed fees and increasing block prices to improve access to piped water.

% This paper demonstrates that pricing policies intended to increase water access especially among the poor can have unintended consequences when low-income households can simply access the piped water network through their neighbors.

% The counterfactual exercises find that a simple two-part tariff structure outperforms the increasing block tariff currently implemented in Manila in terms of overall access to water, total welfare, and even welfare among the bottom half of water users.  In this counterfactual, high fixed costs of accessing an individual connection allow for sharing networks to play an important role in extending water access despite sizable hassle costs.
